---
title: "MMRM vs. Survival analysis for AD clinical trials"
author: "R.G. Thomas"
date: "`r Sys.Date()`"
fontsize: 11pt
geometry: "left=3cm,right=3in,top=2cm,bottom=2cm"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    highlight: tango
    keep_tex: true 
    includes:
      in_header: "~/shr/preamble.tex"
header-includes:
   - \rhead{Lab report}
   - \fancyfoot[L]{\currfilename} %put current file in footer
   - \setlength{\headheight}{13.59999pt}
bibliography: '`r normalizePath("x18.bib")`'
---

<!-- your comment 
\input{notes.tex}
\hspace{4in}
\fbox{
\parbox{3in}{\Large 
Notes: 
\begin{itemize}
\item a \\
\item b  \\
\end{itemize}
}
}
-->

```{r  include=FALSE, echo=F}
rm(list = ls())
wh <- which(search() == "package:nvimcom")
if (wh > 2) {
  for (x in (wh - 1):2) {
    detach(po = x, force = T)
  }
}
rm(wh)
options(
  knitr.kable.NA = "", dplyr.summarise.inform = FALSE,
  dplyr.print_max = 1e9, knitr.table.format = "latex"
)
library(pacman)
p_load(skimr, janitor, kableExtra, tidyverse, knitr, lubridate, readxl)

opts_chunk$set(
  warning = F, message = F, echo = F, fig.width = 3.2,
  fig.height = 3, results = "asis", dev = "pdf",
  opts_chunk$set(cache.path = "cache/")
)
source("~/shr/zz.tools.R")
options(scipen = 1, digits = 2)
```

<!-- your comment 
\listoftodos
\lipsum[2]
\todo[inline, color=cyan]{The original todo note withouth changed colours.\newline Here's another line.}
\begin{wrapfigure}{r}[.4\textwidth]{.4\textwidth}%
\textit{\sffamily\normalsize\textcolor{gray!50}{
Tornado plot: x-axis participants IDs y-axis study visits plot mean of cumulative
AEs stratified by pre-covid and post-covid study visits (March 12 is
the date we are using as a cut-point: through March 12 is pre-COVID,
after March 12 is post-COVID) }}% 
\end{wrapfigure}
\lipsum[2]

Sample reference section reference. @Pocock1975a

\begin{marginfigure}
\includegraphics[width=3in]{/Users/zenn/shr/images/sudoku-3d-cube.pdf} \caption{test}\label{fig:unnamed-chunk-8}
\end{marginfigure}

My most common usage of the todonotes package, is to insert an uncustomized
todonotes somewhere in a latex document. An example of this usage is the command
\todo{Make a cake \ldots},
\lipsum[1]
\todo[color=green]{Make a pie \ldots},
\lipsum[1] 
\todo{Make a joke \ldots},
\listoftodos
-->


# Introduction
Consider the different alternatives in a prospective clinical trial. 
There is incidence, time to event, slope, AUC etc. 

How do analytic methods compare for these difference "estimands."

Consider missing data constructs, MCAR, MAR, MNAR. 

How do analytic methods compare for these difference missing data constructs? 

Start review with Mike D's 2012 paper. 
@donohueMixedModelRepeated2012a

# Methods

# Results
# Code 

<!-- your comment 
```{r, attr.source='.numberLines', ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```
-->

# References
