---
title: "MMRM vs. Survival Analysis for AD Clinical Trials:
  A Historical Review and Simulation Study"
author: "R.G. Thomas"
date: "`r Sys.Date()`"
fontsize: 11pt
geometry: "left=3cm,right=3cm,top=2cm,bottom=2cm"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    highlight: tango
    keep_tex: true
bibliography: references.bib
link-citations: true
---

```{r setup, include=FALSE}
options(
  knitr.kable.NA = "",
  dplyr.summarise.inform = FALSE,
  dplyr.print_max = 1e9,
  knitr.table.format = "latex"
)
library(knitr)
library(kableExtra)
library(tidyverse)
library(survival)

opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.width = 5,
  fig.height = 3.5,
  results = "asis",
  dev = "pdf",
  cache = TRUE,
  cache.path = "cache/"
)
options(scipen = 1, digits = 3)
set.seed(20260310)
```

# Introduction

## The divergent paths of clinical trial analysis

The statistical analysis of clinical trials has followed
markedly different trajectories across therapeutic areas.
In oncology, time-to-event endpoints analyzed by survival
methods remain the regulatory gold standard, with overall
survival serving as the definitive measure of treatment
efficacy [@fda2018oncology; @fleming1996surrogate]. In
contrast, Alzheimer's disease (AD) and other
neurodegenerative disease trials have converged on
change-from-baseline estimands analyzed by the mixed model
for repeated measures (MMRM) as the primary analytic
framework [@donohue2012mmrm; @mallinckrodt2003assessing].
This divergence is not an accident of history but rather
reflects fundamental differences in disease biology,
regulatory philosophy, and the statistical properties of
the available endpoints.

This manuscript traces the historical developments that led
to the current analytic conventions in each field and
presents a simulation study comparing MMRM and survival
analysis applied to the same underlying disease process,
using the Clinical Dementia Rating (CDR) scale as the
primary measure.

## Regulatory origins: the FDA dual-outcome requirement

The regulatory framework for AD clinical trials was shaped
decisively by the 1990 FDA guidelines authored by Paul
Leber, which required demonstration of efficacy on two
co-primary endpoints: one cognitive (such as the
Alzheimer's Disease Assessment Scale-Cognitive Subscale,
ADAS-Cog) and one global or functional measure
[@leber1990guidelines; @rosen1984adas]. The rationale was
that a statistically significant difference on a
psychometric test alone might not reflect clinically
meaningful change; a second, independent measure of global
function was needed to confirm relevance. This dual-outcome
requirement established the precedent that AD trials would
be evaluated on the magnitude and direction of change from
baseline, rather than on the occurrence of discrete clinical
events.

The ADAS-Cog, developed by @rosen1984adas, and the CDR,
codified by @morris1993cdr, became the two pillars of AD
trial assessment. The CDR in particular provided both a
global staging instrument (the global CDR score, ranging
from 0 to 3) and a sum-of-boxes score (CDR-SB, ranging
from 0 to 18) that offered a quasi-continuous measure of
disease severity suitable for change-score analysis. The
2018 FDA guidance on early AD further reinforced the
change-from-baseline paradigm, while allowing greater
flexibility in endpoint selection for earlier disease
stages [@fda2018earlyad].

## The fall of LOCF and the rise of MMRM

Through the 1990s, the standard analytic approach for AD
trials was last observation carried forward (LOCF) combined
with analysis of covariance (ANCOVA) at the final time
point. Under LOCF, a participant's last observed value is
imputed for all subsequent missing visits, and the analysis
proceeds as if no data were missing. This approach was
computationally convenient and widely accepted by
regulators, but it rests on the untenable assumption that
a participant's trajectory would have remained flat from
the time of dropout---an assumption grossly violated in
progressive diseases where dropout is often driven by
worsening symptoms.

The foundational work of @mallinckrodt2001dropout and
@mallinckrodt2001typei demonstrated through simulation
that LOCF produces substantially biased treatment effect
estimates and inflated Type I error rates when dropout is
differential between treatment arms. In a progressive
disease like AD, where placebo-arm participants tend to
drop out at higher rates due to worsening, LOCF
paradoxically imputes favorable values for the placebo
group (freezing them at an earlier, less impaired state),
thereby attenuating the estimated treatment effect.
Conversely, if active-treatment participants drop out due
to adverse events while still relatively well, LOCF
preserves their favorable scores, inflating the apparent
benefit.

@mallinckrodt2003assessing and @mallinckrodt2003response
extended this work by comparing LOCF to the mixed model
for repeated measures (MMRM), which uses all observed data
under a likelihood-based framework that is valid when data
are missing at random (MAR). Unlike LOCF, MMRM does not
impute missing values; instead, it borrows information
from the observed correlation structure across time points
to provide unbiased estimates under MAR. @obrien2005semiparametric
proposed alternative semi-parametric and non-parametric
methods and found that MMRM performed comparably to their
proposed approaches while making less restrictive
assumptions. @molenberghs2004analyzing provided the
theoretical underpinning for likelihood-based analyses of
incomplete longitudinal data.

By the mid-2000s, the weight of evidence against LOCF was
overwhelming, and the 2010 National Research Council panel
report, commissioned by the FDA, formally recommended
against single imputation methods and endorsed
likelihood-based and multiple imputation approaches as
preferred alternatives [@nrc2010missing]. The ICH E9(R1)
addendum on estimands further clarified the conceptual
framework by requiring explicit specification of the
target estimand---including how intercurrent events such
as treatment discontinuation are handled---before
selecting an analytic method [@ich2019e9r1].

## MMRM as the primary analysis in AD

@donohue2012mmrm compared MMRM, slope models, and
quadratic models across five AD Cooperative Study (ADCS)
trials and found that MMRM, which treats time as
categorical and imposes no parametric assumptions on the
mean trajectory, provided robust estimates across a range
of disease stages and trial durations. Although continuous
time (slope) models can be more statistically efficient
when the linearity assumption holds, the unstructured mean
of MMRM offers protection against model misspecification
in a disease where trajectories are heterogeneous and
potentially nonlinear. @wang2018dpm proposed a disease
progression model (DPM) with proportional treatment effects
and demonstrated power gains over MMRM in autosomal
dominant AD, though the DPM requires stronger parametric
assumptions that may limit generalizability.

Today, virtually every pivotal AD trial uses MMRM or a
close variant as the pre-specified primary analysis. The
aducanumab EMERGE and ENGAGE trials
[@buddhaeberlein2022aducanumab] and the lecanemab Study
201 [@dhadda2022lecanemab] all employed MMRM with change
from baseline in CDR-SB as the primary endpoint, with
sensitivity analyses from DPM and other longitudinal
models used to assess robustness.

## Why survival analysis persists in oncology

In oncology, the natural endpoint is death, and overall
survival (OS) has served as the primary measure of
treatment efficacy since the earliest cancer clinical
trials. OS is binary, unambiguous, and universally
meaningful. @fleming1996surrogate cautioned against
replacing OS with surrogate endpoints unless rigorous
validation criteria are met, and the FDA has historically
required OS data for full (as opposed to accelerated)
approval of cancer drugs [@fda2018oncology]. Even when
progression-free survival or response rate serves as the
primary endpoint for accelerated approval, confirmatory
trials typically require OS data.

The persistence of survival analysis in oncology reflects
several features that distinguish it from AD: (1) a clearly
defined, irreversible event (death); (2) a well-understood
proportional hazards framework with established regulatory
precedent; (3) a tradition of interim analyses based on
information fractions from event counts; and (4) the
absence of a continuous, validated severity measure that
changes monotonically and is universally accepted across
tumor types.

## Rate of change vs. time to event in AD

Given that CDR and similar scales can be used to define
both change scores (amenable to MMRM) and threshold
crossings (amenable to survival analysis), the question
arises: what is the relative efficiency of these two
approaches when applied to the same underlying data?

@donohue2011relative derived an analytic inflation factor
comparing Cox proportional hazards models of time to a
threshold with linear models of continuous outcomes, and
demonstrated through simulation that linear models
consistently yielded greater statistical power for
detecting treatment effects in ADNI data. @li2019relative
extended this work to presymptomatic AD, finding that
power was approximately doubled with repeated continuous
measures compared to time-to-progression analysis.
@jonsson2024progression introduced a progression model for
repeated measures (PMRM) that expresses treatment effects
in time units (days of delay) rather than CDR-SB points,
offering a clinically interpretable alternative while
maintaining efficiency comparable to MMRM.
@ito2014predicting modeled CDR-SB trajectories to predict
time to clinically meaningful worsening in MCI patients,
finding that the median time to a 1-point change was
approximately 2 years---informing both trial duration and
sample size planning.

## Present study

Despite the theoretical and empirical evidence favoring
continuous measures, the comparison between MMRM and
survival analysis for AD endpoints has not been
systematically examined in a simulation that uses the
global CDR as both a continuous outcome and the basis for
a survival endpoint defined by conversion from CDR 0.5
(MCI/very mild dementia) to CDR 1.0 or higher (mild
dementia). The present study addresses this gap through a
Monte Carlo simulation that generates realistic CDR
trajectories under a plausible data-generating model and
compares the statistical properties of MMRM and Cox
regression applied to the same simulated trial data.

# Methods

## Data generating model

We simulate a two-arm, placebo-controlled trial with
$n = 200$ participants per arm, all entering with a
baseline global CDR of 0.5. The primary continuous outcome
is the global CDR score observed at visits scheduled every
6 months over 24 months (5 post-baseline visits). The
survival endpoint is time to conversion from CDR 0.5 to
CDR $\geq$ 1.0.

The latent continuous disease severity $Y_{ij}^*$ for
participant $i$ at visit $j$ is modeled as:

$$
Y_{ij}^* = \beta_0 + \beta_1 t_j + \beta_2 T_i
  + \beta_3 (t_j \times T_i) + b_{0i} + b_{1i} t_j
  + \epsilon_{ij}
$$

where $t_j$ is visit time in years, $T_i$ is the
treatment indicator (0 = placebo, 1 = active), $b_{0i}$
and $b_{1i}$ are participant-specific random intercepts
and slopes drawn from a bivariate normal distribution,
and $\epsilon_{ij} \sim N(0, \sigma^2)$ is residual error.

The observed global CDR is obtained by mapping $Y^*$ to
the CDR staging categories {0, 0.5, 1, 2, 3} using
threshold parameters calibrated to observed CDR
distributions in ADNI.

## Parameter values

The simulation parameters are selected to reflect
plausible progression rates in a population with MCI due
to AD (CDR 0.5 at baseline):

```{r sim-params}
params <- tibble::tribble(
  ~Parameter, ~Value, ~Description,
  "$\\beta_0$", "0.5", "Baseline CDR (MCI)",
  "$\\beta_1$", "0.30", "Annual rate, placebo (CDR/yr)",
  "$\\beta_2$", "0.00", "Baseline treatment effect",
  "$\\beta_3$", "-0.075",
    "Treatment x time (25\\% slowing)",
  "$\\sigma_b^2$ (intercept)", "0.04",
    "Random intercept variance",
  "$\\sigma_b^2$ (slope)", "0.02",
    "Random slope variance",
  "$\\rho$", "0.3",
    "Intercept-slope correlation",
  "$\\sigma^2$", "0.01",
    "Residual variance",
  "Dropout rate", "15\\%/year",
    "MAR dropout, severity-dependent"
)
kable(
  params,
  booktabs = TRUE,
  escape = FALSE,
  caption = "Simulation parameter values."
) |>
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 9
  )
```

## Design specifications

- **Sample size:** 200 per arm (400 total)
- **Visit schedule:** Baseline, 6, 12, 18, 24 months
- **Survival endpoint:** Time from randomization to first
  visit with global CDR $\geq$ 1.0
- **MMRM endpoint:** Change from baseline in latent
  CDR score at 24 months
- **Treatment effect:** 25% slowing of progression
  (hazard ratio $\approx$ 0.75 for the survival endpoint)

## Simulation procedure

```{r sim-function, echo=TRUE, eval=FALSE}
sim_trial <- function(n_per_arm = 200,
                      n_visits = 5,
                      beta = c(0.5, 0.30, 0, -0.075),
                      sigma_b = matrix(
                        c(0.04, 0.3 * sqrt(0.04 * 0.02),
                          0.3 * sqrt(0.04 * 0.02), 0.02),
                        2, 2
                      ),
                      sigma_e = 0.1,
                      dropout_rate = 0.15) {
  n <- 2 * n_per_arm
  trt <- rep(0:1, each = n_per_arm)
  times <- seq(0.5, 2.5, by = 0.5)

  re <- MASS::mvrnorm(n, mu = c(0, 0), Sigma = sigma_b)

  records <- list()
  for (i in seq_len(n)) {
    y_star <- numeric(n_visits)
    dropped <- FALSE
    for (j in seq_len(n_visits)) {
      if (dropped) next
      t_j <- times[j]
      mu_ij <- beta[1] + beta[2] * t_j +
        beta[3] * trt[i] + beta[4] * (t_j * trt[i]) +
        re[i, 1] + re[i, 2] * t_j
      y_star[j] <- mu_ij + rnorm(1, 0, sigma_e)
      cdr_obs <- cut(
        y_star[j],
        breaks = c(-Inf, 0.25, 0.75, 1.5, 2.5, Inf),
        labels = c(0, 0.5, 1, 2, 3)
      )
      cdr_obs <- as.numeric(as.character(cdr_obs))
      p_drop <- 1 - exp(-dropout_rate * 0.5 *
        (1 + 0.5 * (y_star[j] - 0.5)))
      if (runif(1) < p_drop) dropped <- TRUE
      records[[length(records) + 1]] <- data.frame(
        id = i, trt = trt[i], visit = j,
        time = t_j, y_star = y_star[j],
        cdr = cdr_obs
      )
    }
  }
  do.call(rbind, records)
}
```

For each of $R = 2000$ replications, we:

1. Generate a complete trial dataset using the model above.
2. Apply severity-dependent MAR dropout.
3. Fit an MMRM to the change from baseline in the latent
   CDR score, with treatment, visit, treatment-by-visit
   interaction, and baseline score as covariates, using an
   unstructured covariance matrix.
4. Fit a Cox proportional hazards model to time to first
   CDR $\geq$ 1.0, with treatment as the sole covariate.
5. Record the treatment effect estimate, standard error,
   p-value, and 95% confidence interval from each model.

## Analysis plan

```{r analysis-functions, echo=TRUE, eval=FALSE}
fit_mmrm <- function(dat) {
  dat <- dat |>
    group_by(id) |>
    mutate(
      baseline = first(y_star),
      change = y_star - baseline
    ) |>
    ungroup() |>
    filter(visit > 0) |>
    mutate(visit_f = factor(visit))

  mod <- nlme::gls(
    change ~ trt * visit_f + baseline,
    data = dat,
    correlation = nlme::corSymm(form = ~ visit | id),
    weights = nlme::varIdent(form = ~ 1 | visit_f),
    na.action = na.omit
  )
  trt_24 <- summary(mod)$tTable["trt", ]
  c(est = trt_24[1], se = trt_24[2], p = trt_24[4])
}

fit_cox <- function(dat) {
  surv_dat <- dat |>
    group_by(id) |>
    summarise(
      trt = first(trt),
      event_time = {
        conv <- which(cdr >= 1)
        if (length(conv) > 0) time[conv[1]] else max(time)
      },
      event = as.numeric(any(cdr >= 1)),
      .groups = "drop"
    )
  mod <- coxph(
    Surv(event_time, event) ~ trt,
    data = surv_dat
  )
  s <- summary(mod)
  c(
    est = coef(mod),
    se = s$coefficients[, "se(coef)"],
    p = s$coefficients[, "Pr(>|z|)"]
  )
}
```

## Performance metrics

For each method, we compute:

- **Bias:** Mean estimate minus true parameter value
- **Empirical SE:** Standard deviation of estimates across
  replications
- **Mean model-based SE:** Mean of estimated standard errors
- **Power:** Proportion of replications with $p < 0.05$
- **Coverage:** Proportion of 95% CIs containing the true
  value
- **Convergence rate:** Proportion of replications
  producing valid estimates

# Results

```{r run-simulation, cache=TRUE, eval=FALSE}
n_reps <- 2000
results <- vector("list", n_reps)

for (r in seq_len(n_reps)) {
  dat <- sim_trial()
  mmrm_res <- tryCatch(
    fit_mmrm(dat), error = function(e) rep(NA, 3)
  )
  cox_res <- tryCatch(
    fit_cox(dat), error = function(e) rep(NA, 3)
  )
  results[[r]] <- data.frame(
    rep = r,
    mmrm_est = mmrm_res[1],
    mmrm_se = mmrm_res[2],
    mmrm_p = mmrm_res[3],
    cox_est = cox_res[1],
    cox_se = cox_res[2],
    cox_p = cox_res[3]
  )
}

sim_results <- do.call(rbind, results)
```

## Summary statistics

*[Results table to be populated after simulation execution.]*

```{r summary-table, eval=FALSE}
metrics <- sim_results |>
  summarise(
    mmrm_bias = mean(mmrm_est, na.rm = TRUE) -
      (-0.075 * 2),
    mmrm_emp_se = sd(mmrm_est, na.rm = TRUE),
    mmrm_mean_se = mean(mmrm_se, na.rm = TRUE),
    mmrm_power = mean(mmrm_p < 0.05, na.rm = TRUE),
    cox_bias = mean(cox_est, na.rm = TRUE) -
      log(0.75),
    cox_emp_se = sd(cox_est, na.rm = TRUE),
    cox_mean_se = mean(cox_se, na.rm = TRUE),
    cox_power = mean(cox_p < 0.05, na.rm = TRUE)
  )

summary_tab <- tibble::tribble(
  ~Metric, ~MMRM, ~Cox,
  "Bias", metrics$mmrm_bias, metrics$cox_bias,
  "Empirical SE", metrics$mmrm_emp_se,
    metrics$cox_emp_se,
  "Mean model SE", metrics$mmrm_mean_se,
    metrics$cox_mean_se,
  "Power", metrics$mmrm_power, metrics$cox_power
)

kable(
  summary_tab,
  booktabs = TRUE,
  digits = 3,
  caption = "Simulation results: MMRM vs. Cox regression
    (2000 replications)."
) |>
  kable_styling(latex_options = "hold_position")
```

## Figures

```{r power-comparison, eval=FALSE, fig.cap="Power comparison across sample sizes."}
sample_sizes <- c(100, 150, 200, 300, 400)
power_results <- map_dfr(sample_sizes, function(n) {
  # placeholder for power across sample sizes
  tibble(
    n = n,
    method = c("MMRM", "Cox PH"),
    power = c(NA_real_, NA_real_)
  )
})

ggplot(power_results, aes(x = n, y = power,
    linetype = method, shape = method)) +
  geom_line() +
  geom_point(size = 2) +
  scale_linetype_manual(
    values = c("solid", "dashed")
  ) +
  scale_shape_manual(values = c(16, 17)) +
  labs(
    x = "Sample size per arm",
    y = "Power",
    linetype = "Method",
    shape = "Method"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

```{r density-plot, eval=FALSE, fig.cap="Distribution of treatment effect estimates across replications."}
sim_long <- sim_results |>
  select(rep, mmrm_est, cox_est) |>
  pivot_longer(
    cols = c(mmrm_est, cox_est),
    names_to = "method",
    values_to = "estimate"
  ) |>
  mutate(
    method = ifelse(
      method == "mmrm_est", "MMRM", "Cox PH"
    )
  )

ggplot(sim_long, aes(x = estimate,
    linetype = method)) +
  geom_density(fill = "grey85", alpha = 0.5) +
  labs(
    x = "Treatment effect estimate",
    y = "Density",
    linetype = "Method"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

# Discussion

The historical review reveals that the convergence of AD
trials on MMRM and change-score estimands resulted from
three reinforcing forces: (1) the regulatory requirement
for continuous cognitive and functional co-primary
endpoints, established by @leber1990guidelines; (2) the
decisive empirical evidence against LOCF as a primary
analysis method, led by @mallinckrodt2001dropout and
codified by the @nrc2010missing panel; and (3) the
statistical efficiency advantages of modeling repeated
continuous data relative to dichotomizing into time-to-event
endpoints, as demonstrated by @donohue2011relative and
@li2019relative.

Oncology, by contrast, arrived at a different equilibrium
because its primary endpoint---death---is inherently
binary and irreversible, because the proportional hazards
framework has deep regulatory precedent, and because no
single continuous severity measure plays the role that
ADAS-Cog or CDR-SB plays in AD
[@fleming1996surrogate; @fda2018oncology].

The simulation results are expected to confirm the
theoretical predictions from the literature: MMRM should
yield substantially greater power than Cox regression when
both are applied to CDR data from the same simulated
trial, because the continuous analysis uses information
from the full trajectory of scores at every visit, whereas
the survival analysis reduces this trajectory to a single
binary event. The power advantage should be most pronounced
in the MCI-to-mild-dementia transition modeled here, where
conversion rates are modest and many participants in a
24-month trial will not reach the CDR $\geq$ 1.0 threshold,
resulting in heavy right-censoring that limits the
information available to the Cox model.

Several caveats warrant emphasis. First, the comparison is
inherently asymmetric: the MMRM uses more information from
each participant (all visits) than the Cox model (only the
first conversion event). Second, the survival endpoint is
limited by the visit schedule; true conversion may occur
between visits, introducing interval censoring that the
standard Cox model does not accommodate. Third, the
relative efficiency depends on the threshold chosen; a
threshold that most participants cross during follow-up
would shift the balance toward the survival analysis.
Fourth, the clinical relevance of the two estimands
differs: a treatment effect expressed as slowing of
CDR-SB decline addresses a different clinical question than
one expressed as delay to a diagnostic milestone. Both may
be meaningful to patients and regulators, and the choice
between them should be driven by the estimand framework of
@ich2019e9r1 rather than by statistical convenience alone.

# Future research

1. **Additional covariance structures.** The present
   simulation assumes MAR dropout. Extensions under MNAR
   mechanisms (e.g., pattern-mixture models, selection
   models) would illuminate the robustness of each method
   under more adversarial missingness.

2. **Nonlinear trajectories.** AD progression is not
   strictly linear, particularly in the transition from
   MCI to dementia. Simulation of sigmoidal or piecewise
   trajectories would test the sensitivity of MMRM (which
   makes no linearity assumption) relative to slope-based
   alternatives.

3. **Alternative survival formulations.** Interval-censored
   survival models and multi-state models that accommodate
   transitions among CDR stages (0.5 $\to$ 1 $\to$ 2
   $\to$ 3) could recover some of the information lost by
   standard Cox regression.

4. **Joint modeling.** Joint models of longitudinal CDR
   trajectories and the time-to-conversion event could
   unify the two analytic approaches, potentially offering
   both the efficiency of longitudinal modeling and the
   clinical interpretability of a time-to-event estimand.

5. **Software tools.** An R package implementing the
   simulation framework and providing power calculation
   utilities for both MMRM and survival designs in AD
   trials would facilitate trial planning.

6. **Applied validation.** Retrospective application of
   both methods to completed trial datasets (e.g., from the
   ADCS or ADNI) would ground-truth the simulation
   findings in real data with genuine missingness patterns
   and measurement properties.

7. **Estimand alignment.** Further work is needed to
   clarify how the ICH E9(R1) estimand framework should
   be applied when the same underlying data support both
   change-score and time-to-event estimands, particularly
   regarding the handling of intercurrent events such as
   treatment switching or rescue medication.

# References
