% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\documentclass[
  11pt,
]{article}
\usepackage{xcolor}
\usepackage[left=3cm,right=5cm,top=2cm,bottom=2cm]{geometry}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={MMRM vs.~Survival Analysis for AD Clinical Trials: A Historical Review and Simulation Study},
  pdfauthor={R.G. Thomas},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{MMRM vs.~Survival Analysis for AD Clinical Trials: A Historical
Review and Simulation Study}
\author{R.G. Thomas}
\date{April 19, 2026}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\section{Introduction}\label{introduction}

\subsection{The divergent paths of clinical trial
analysis}\label{the-divergent-paths-of-clinical-trial-analysis}

The statistical analysis of clinical trials has followed markedly
different trajectories across therapeutic areas. In oncology,
time-to-event endpoints analyzed by survival methods remain the
regulatory gold standard, with overall survival serving as the
definitive measure of treatment
efficacy\textsuperscript{\citeproc{ref-fda2018oncology}{1},\citeproc{ref-fleming1996surrogate}{2}}.
In contrast, Alzheimer's disease (AD) and other neurodegenerative
disease trials have converged on change-from-baseline estimands analyzed
by the mixed model for repeated measures (MMRM) as the primary analytic
framework\textsuperscript{\citeproc{ref-donohue2012mmrm}{3},\citeproc{ref-mallinckrodt2003assessing}{4}}.
This divergence is not an accident of history but rather reflects
fundamental differences in disease biology, regulatory philosophy, and
the statistical properties of the available endpoints.

This manuscript traces the historical developments that led to the
current analytic conventions in each field and presents a simulation
study comparing MMRM and survival analysis applied to the same
underlying disease process, using the Clinical Dementia Rating (CDR)
scale as the primary measure.

\subsection{Regulatory origins: the FDA dual-outcome
requirement}\label{regulatory-origins-the-fda-dual-outcome-requirement}

The regulatory framework for AD clinical trials was shaped decisively by
the 1990 FDA guidelines authored by Paul Leber, which required
demonstration of efficacy on two co-primary endpoints: one cognitive
(such as the Alzheimer's Disease Assessment Scale-Cognitive Subscale,
ADAS-Cog) and one global or functional
measure\textsuperscript{\citeproc{ref-leber1990guidelines}{5},\citeproc{ref-rosen1984adas}{6}}.
The rationale was that a statistically significant difference on a
psychometric test alone might not reflect clinically meaningful change;
a second, independent measure of global function was needed to confirm
relevance. This dual-outcome requirement established the precedent that
AD trials would be evaluated on the magnitude and direction of change
from baseline, rather than on the occurrence of discrete clinical
events.

The ADAS-Cog, developed
by\textsuperscript{\citeproc{ref-rosen1984adas}{6}}, and the CDR,
codified by\textsuperscript{\citeproc{ref-morris1993cdr}{7}}, became the
two pillars of AD trial assessment. The CDR in particular provided both
a global staging instrument (the global CDR score, ranging from 0 to 3)
and a sum-of-boxes score (CDR-SB, ranging from 0 to 18) that offered a
quasi-continuous measure of disease severity suitable for change-score
analysis. The 2018 FDA guidance on early AD further reinforced the
change-from-baseline paradigm, while allowing greater flexibility in
endpoint selection for earlier disease
stages\textsuperscript{\citeproc{ref-fda2018earlyad}{8}}.

\subsection{The fall of LOCF and the rise of
MMRM}\label{the-fall-of-locf-and-the-rise-of-mmrm}

Through the 1990s, the standard analytic approach for AD trials was last
observation carried forward (LOCF) combined with analysis of covariance
(ANCOVA) at the final time point. Under LOCF, a participant's last
observed value is imputed for all subsequent missing visits, and the
analysis proceeds as if no data were missing. This approach was
computationally convenient and widely accepted by regulators, but it
rests on the untenable assumption that a participant's trajectory would
have remained flat from the time of dropout---an assumption grossly
violated in progressive diseases where dropout is often driven by
worsening symptoms.

The foundational work
of\textsuperscript{\citeproc{ref-mallinckrodt2001dropout}{9}}
and\textsuperscript{\citeproc{ref-mallinckrodt2001typei}{10}}
demonstrated through simulation that LOCF produces substantially biased
treatment effect estimates and inflated Type I error rates when dropout
is differential between treatment arms. In a progressive disease like
AD, where placebo-arm participants tend to drop out at higher rates due
to worsening, LOCF paradoxically imputes favorable values for the
placebo group (freezing them at an earlier, less impaired state),
thereby attenuating the estimated treatment effect. Conversely, if
active-treatment participants drop out due to adverse events while still
relatively well, LOCF preserves their favorable scores, inflating the
apparent benefit.

\textsuperscript{\citeproc{ref-mallinckrodt2003assessing}{4}}
and\textsuperscript{\citeproc{ref-mallinckrodt2003response}{11}}
extended this work by comparing LOCF to the mixed model for repeated
measures (MMRM), which uses all observed data under a likelihood-based
framework that is valid when data are missing at random (MAR). Unlike
LOCF, MMRM does not impute missing values; instead, it borrows
information from the observed correlation structure across time points
to provide unbiased estimates under
MAR.\textsuperscript{\citeproc{ref-obrien2005semiparametric}{12}}
proposed alternative semi-parametric and non-parametric methods and
found that MMRM performed comparably to their proposed approaches while
making less restrictive
assumptions.\textsuperscript{\citeproc{ref-molenberghs2004analyzing}{13}}
provided the theoretical underpinning for likelihood-based analyses of
incomplete longitudinal data.

By the mid-2000s, the weight of evidence against LOCF was overwhelming,
and the 2010 National Research Council panel report, commissioned by the
FDA, formally recommended against single imputation methods and endorsed
likelihood-based and multiple imputation approaches as preferred
alternatives\textsuperscript{\citeproc{ref-nrc2010missing}{14}}. The ICH
E9(R1) addendum on estimands further clarified the conceptual framework
by requiring explicit specification of the target estimand---including
how intercurrent events such as treatment discontinuation are
handled---before selecting an analytic
method\textsuperscript{\citeproc{ref-ich2019e9r1}{15}}.

\subsection{MMRM as the primary analysis in
AD}\label{mmrm-as-the-primary-analysis-in-ad}

\textsuperscript{\citeproc{ref-donohue2012mmrm}{3}} compared MMRM, slope
models, and quadratic models across five AD Cooperative Study (ADCS)
trials and found that MMRM, which treats time as categorical and imposes
no parametric assumptions on the mean trajectory, provided robust
estimates across a range of disease stages and trial durations. Although
continuous time (slope) models can be more statistically efficient when
the linearity assumption holds, the unstructured mean of MMRM offers
protection against model misspecification in a disease where
trajectories are heterogeneous and potentially
nonlinear.\textsuperscript{\citeproc{ref-wang2018dpm}{16}} proposed a
disease progression model (DPM) with proportional treatment effects and
demonstrated power gains over MMRM in autosomal dominant AD, though the
DPM requires stronger parametric assumptions that may limit
generalizability.

Today, virtually every pivotal AD trial uses MMRM or a close variant as
the pre-specified primary analysis. The aducanumab EMERGE and ENGAGE
trials\textsuperscript{\citeproc{ref-buddhaeberlein2022aducanumab}{17}}
and the lecanemab Study
201\textsuperscript{\citeproc{ref-dhadda2022lecanemab}{18}} all employed
MMRM with change from baseline in CDR-SB as the primary endpoint, with
sensitivity analyses from DPM and other longitudinal models used to
assess robustness.

\subsection{Why survival analysis persists in
oncology}\label{why-survival-analysis-persists-in-oncology}

In oncology, the natural endpoint is death, and overall survival (OS)
has served as the primary measure of treatment efficacy since the
earliest cancer clinical trials. OS is binary, unambiguous, and
universally
meaningful.\textsuperscript{\citeproc{ref-fleming1996surrogate}{2}}
cautioned against replacing OS with surrogate endpoints unless rigorous
validation criteria are met, and the FDA has historically required OS
data for full (as opposed to accelerated) approval of cancer
drugs\textsuperscript{\citeproc{ref-fda2018oncology}{1}}. Even when
progression-free survival or response rate serves as the primary
endpoint for accelerated approval, confirmatory trials typically require
OS data.

The persistence of survival analysis in oncology reflects several
features that distinguish it from AD: (1) a clearly defined,
irreversible event (death); (2) a well-understood proportional hazards
framework with established regulatory precedent; (3) a tradition of
interim analyses based on information fractions from event counts; and
(4) the absence of a continuous, validated severity measure that changes
monotonically and is universally accepted across tumor types.

\subsection{Rate of change vs.~time to event in
AD}\label{rate-of-change-vs.-time-to-event-in-ad}

Given that CDR and similar scales can be used to define both change
scores (amenable to MMRM) and threshold crossings (amenable to survival
analysis), the question arises: what is the relative efficiency of these
two approaches when applied to the same underlying data?

\textsuperscript{\citeproc{ref-donohue2011relative}{19}} derived an
analytic inflation factor comparing Cox proportional hazards models of
time to a threshold with linear models of continuous outcomes, and
demonstrated through simulation that linear models consistently yielded
greater statistical power for detecting treatment effects in ADNI
data.\textsuperscript{\citeproc{ref-li2019relative}{20}} extended this
work to presymptomatic AD, finding that power was approximately doubled
with repeated continuous measures compared to time-to-progression
analysis.\textsuperscript{\citeproc{ref-jonsson2024progression}{21}}
introduced a progression model for repeated measures (PMRM) that
expresses treatment effects in time units (days of delay) rather than
CDR-SB points, offering a clinically interpretable alternative while
maintaining efficiency comparable to
MMRM.\textsuperscript{\citeproc{ref-ito2014predicting}{22}} modeled
CDR-SB trajectories to predict time to clinically meaningful worsening
in MCI patients, finding that the median time to a 1-point change was
approximately 2 years---informing both trial duration and sample size
planning.

\subsection{Present study}\label{present-study}

Despite the theoretical and empirical evidence favoring continuous
measures, the comparison between MMRM and survival analysis for AD
endpoints has not been systematically examined in a simulation that uses
the global CDR as both a continuous outcome and the basis for a survival
endpoint defined by conversion from CDR 0.5 (MCI/very mild dementia) to
CDR 1.0 or higher (mild dementia). The present study addresses this gap
through a Monte Carlo simulation that generates realistic CDR
trajectories under a plausible data-generating model and compares the
statistical properties of MMRM and Cox regression applied to the same
simulated trial data.

\section{Methods}\label{methods}

\subsection{ADEMP structure}\label{ademp-structure}

Following Morris, White, and
Crowther\textsuperscript{\citeproc{ref-morris2019using}{23}}, the
simulation is reported under the ADEMP framework.

\begin{itemize}
\tightlist
\item
  \textbf{Aims.} Compare the operating characteristics of an MMRM on
  change-from-baseline in latent CDR versus a Cox proportional hazards
  model on time to first CDR \(\geq 1.0\), under an Alzheimer's-disease
  DGM with informative dropout.
\item
  \textbf{Data-generating mechanism.} Mixed-effects model for latent CDR
  trajectories with a discretized observed CDR and outcome-dependent
  dropout; details below.
\item
  \textbf{Estimands.}

  \begin{enumerate}
  \def\labelenumi{(\roman{enumi})}
  \tightlist
  \item
    For MMRM: the treatment effect on change from baseline in latent CDR
    at the LAST visit, expressed as the linear combination
    \(\beta_{\mathrm{trt}} + \beta_{\mathrm{trt:visit}_K}\).
  \item
    For Cox: the log-hazard ratio for time to first CDR \(\geq 1.0\).
  \end{enumerate}
\item
  \textbf{Methods.} MMRM via \texttt{nlme::gls} with unstructured
  covariance; Cox via \texttt{survival::coxph}.
\item
  \textbf{Performance measures.} Bias, empirical SE, mean model SE,
  power, coverage, and convergence rate, each reported with its Monte
  Carlo SE per Morris Table 6. With \(n_{\mathrm{sim}} = 2000\), the
  Monte Carlo SE for power at \(p = 0.5\) is
  \(\sqrt{0.5 \times 0.5 / 2000} = 0.011\); for coverage at \(0.95\),
  MCSE \(\approx 0.005\).
\end{itemize}

\subsection{Data generating model}\label{data-generating-model}

We simulate a two-arm, placebo-controlled trial with \(n = 200\)
participants per arm, all entering with a baseline global CDR of 0.5.
The primary continuous outcome is the global CDR score observed at
visits scheduled every 6 months over 24 months (5 post-baseline visits).
The survival endpoint is time to conversion from CDR 0.5 to CDR \(\geq\)
1.0.

The latent continuous disease severity \(Y_{ij}^*\) for participant
\(i\) at visit \(j\) is modeled as:

\[
Y_{ij}^* = \beta_0 + \beta_1 t_j + \beta_2 T_i
  + \beta_3 (t_j \times T_i) + b_{0i} + b_{1i} t_j
  + \epsilon_{ij}
\]

where \(t_j\) is visit time in years, \(T_i\) is the treatment indicator
(0 = placebo, 1 = active), \(b_{0i}\) and \(b_{1i}\) are
participant-specific random intercepts and slopes drawn from a bivariate
normal distribution, and \(\epsilon_{ij} \sim N(0, \sigma^2)\) is
residual error.

The observed global CDR is obtained by mapping \(Y^*\) to the CDR
staging categories \{0, 0.5, 1, 2, 3\} using threshold parameters
calibrated to observed CDR distributions in ADNI.

\subsection{Parameter values}\label{parameter-values}

The simulation parameters are selected to reflect plausible progression
rates in a population with MCI due to AD (CDR 0.5 at baseline):

\begin{table}[!h]
\centering
\caption{\label{tab:sim-params}Simulation parameter values.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{lll}
\toprule
Parameter & Value & Description\\
\midrule
$\beta_0$ & 0.5 & Baseline CDR (MCI)\\
$\beta_1$ & 0.30 & Annual rate, placebo (CDR/yr)\\
$\beta_2$ & 0.00 & Baseline treatment effect\\
$\beta_3$ & -0.075 & Treatment x time (25\% slowing)\\
$\sigma_b^2$ (intercept) & 0.04 & Random intercept variance\\
\addlinespace
$\sigma_b^2$ (slope) & 0.02 & Random slope variance\\
$\rho$ & 0.3 & Intercept-slope correlation\\
$\sigma^2$ & 0.01 & Residual variance\\
Dropout rate & 15\%/year & MAR dropout, severity-dependent\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Design specifications}\label{design-specifications}

\begin{itemize}
\tightlist
\item
  \textbf{Sample size:} 200 per arm (400 total)
\item
  \textbf{Visit schedule:} Baseline, 6, 12, 18, 24 months
\item
  \textbf{Survival endpoint:} Time from randomization to first visit
  with global CDR \(\geq\) 1.0
\item
  \textbf{MMRM endpoint:} Change from baseline in latent CDR score at 24
  months
\item
  \textbf{Treatment effect:} 25\% slowing of progression (hazard ratio
  \(\approx\) 0.75 for the survival endpoint)
\end{itemize}

\subsection{Simulation procedure}\label{simulation-procedure}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sim\_trial }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(}\AttributeTok{n\_per\_arm =} \DecValTok{200}\NormalTok{,}
                      \AttributeTok{n\_visits =} \DecValTok{5}\NormalTok{,}
                      \AttributeTok{beta =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.5}\NormalTok{, }\FloatTok{0.30}\NormalTok{, }\DecValTok{0}\NormalTok{, }\SpecialCharTok{{-}}\FloatTok{0.075}\NormalTok{),}
                      \AttributeTok{sigma\_b =} \FunctionTok{matrix}\NormalTok{(}
                        \FunctionTok{c}\NormalTok{(}\FloatTok{0.04}\NormalTok{, }\FloatTok{0.3} \SpecialCharTok{*} \FunctionTok{sqrt}\NormalTok{(}\FloatTok{0.04} \SpecialCharTok{*} \FloatTok{0.02}\NormalTok{),}
                          \FloatTok{0.3} \SpecialCharTok{*} \FunctionTok{sqrt}\NormalTok{(}\FloatTok{0.04} \SpecialCharTok{*} \FloatTok{0.02}\NormalTok{), }\FloatTok{0.02}\NormalTok{),}
                        \DecValTok{2}\NormalTok{, }\DecValTok{2}
\NormalTok{                      ),}
                      \AttributeTok{sigma\_e =} \FloatTok{0.1}\NormalTok{,}
                      \AttributeTok{dropout\_rate =} \FloatTok{0.15}\NormalTok{) \{}
\NormalTok{  n }\OtherTok{\textless{}{-}} \DecValTok{2} \SpecialCharTok{*}\NormalTok{ n\_per\_arm}
\NormalTok{  trt }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{0}\SpecialCharTok{:}\DecValTok{1}\NormalTok{, }\AttributeTok{each =}\NormalTok{ n\_per\_arm)}
\NormalTok{  times }\OtherTok{\textless{}{-}} \FunctionTok{seq}\NormalTok{(}\FloatTok{0.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{, }\AttributeTok{by =} \FloatTok{0.5}\NormalTok{)}

\NormalTok{  re }\OtherTok{\textless{}{-}}\NormalTok{ MASS}\SpecialCharTok{::}\FunctionTok{mvrnorm}\NormalTok{(n, }\AttributeTok{mu =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{), }\AttributeTok{Sigma =}\NormalTok{ sigma\_b)}

\NormalTok{  records }\OtherTok{\textless{}{-}} \FunctionTok{list}\NormalTok{()}
  \ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \FunctionTok{seq\_len}\NormalTok{(n)) \{}
\NormalTok{    y\_star }\OtherTok{\textless{}{-}} \FunctionTok{numeric}\NormalTok{(n\_visits)}
\NormalTok{    dropped }\OtherTok{\textless{}{-}} \ConstantTok{FALSE}
    \ControlFlowTok{for}\NormalTok{ (j }\ControlFlowTok{in} \FunctionTok{seq\_len}\NormalTok{(n\_visits)) \{}
      \ControlFlowTok{if}\NormalTok{ (dropped) }\ControlFlowTok{next}
\NormalTok{      t\_j }\OtherTok{\textless{}{-}}\NormalTok{ times[j]}
\NormalTok{      mu\_ij }\OtherTok{\textless{}{-}}\NormalTok{ beta[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{+}\NormalTok{ beta[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ t\_j }\SpecialCharTok{+}
\NormalTok{        beta[}\DecValTok{3}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ trt[i] }\SpecialCharTok{+}\NormalTok{ beta[}\DecValTok{4}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ (t\_j }\SpecialCharTok{*}\NormalTok{ trt[i]) }\SpecialCharTok{+}
\NormalTok{        re[i, }\DecValTok{1}\NormalTok{] }\SpecialCharTok{+}\NormalTok{ re[i, }\DecValTok{2}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ t\_j}
\NormalTok{      y\_star[j] }\OtherTok{\textless{}{-}}\NormalTok{ mu\_ij }\SpecialCharTok{+} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{, sigma\_e)}
\NormalTok{      cdr\_obs }\OtherTok{\textless{}{-}} \FunctionTok{cut}\NormalTok{(}
\NormalTok{        y\_star[j],}
        \AttributeTok{breaks =} \FunctionTok{c}\NormalTok{(}\SpecialCharTok{{-}}\ConstantTok{Inf}\NormalTok{, }\FloatTok{0.25}\NormalTok{, }\FloatTok{0.75}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{, }\ConstantTok{Inf}\NormalTok{),}
        \AttributeTok{labels =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\FloatTok{0.5}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{3}\NormalTok{)}
\NormalTok{      )}
\NormalTok{      cdr\_obs }\OtherTok{\textless{}{-}} \FunctionTok{as.numeric}\NormalTok{(}\FunctionTok{as.character}\NormalTok{(cdr\_obs))}
\NormalTok{      p\_drop }\OtherTok{\textless{}{-}} \DecValTok{1} \SpecialCharTok{{-}} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{dropout\_rate }\SpecialCharTok{*} \FloatTok{0.5} \SpecialCharTok{*}
\NormalTok{        (}\DecValTok{1} \SpecialCharTok{+} \FloatTok{0.5} \SpecialCharTok{*}\NormalTok{ (y\_star[j] }\SpecialCharTok{{-}} \FloatTok{0.5}\NormalTok{)))}
      \ControlFlowTok{if}\NormalTok{ (}\FunctionTok{runif}\NormalTok{(}\DecValTok{1}\NormalTok{) }\SpecialCharTok{\textless{}}\NormalTok{ p\_drop) dropped }\OtherTok{\textless{}{-}} \ConstantTok{TRUE}
\NormalTok{      records[[}\FunctionTok{length}\NormalTok{(records) }\SpecialCharTok{+} \DecValTok{1}\NormalTok{]] }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}
        \AttributeTok{id =}\NormalTok{ i, }\AttributeTok{trt =}\NormalTok{ trt[i], }\AttributeTok{visit =}\NormalTok{ j,}
        \AttributeTok{time =}\NormalTok{ t\_j, }\AttributeTok{y\_star =}\NormalTok{ y\_star[j],}
        \AttributeTok{cdr =}\NormalTok{ cdr\_obs}
\NormalTok{      )}
\NormalTok{    \}}
\NormalTok{  \}}
  \FunctionTok{do.call}\NormalTok{(rbind, records)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

For each of \(R = 2000\) replications, we:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Generate a complete trial dataset using the model above.
\item
  Apply severity-dependent MAR dropout.
\item
  Fit an MMRM to the change from baseline in the latent CDR score, with
  treatment, visit, treatment-by-visit interaction, and baseline score
  as covariates, using an unstructured covariance matrix.
\item
  Fit a Cox proportional hazards model to time to first CDR \(\geq\)
  1.0, with treatment as the sole covariate.
\item
  Record the treatment effect estimate, standard error, p-value, and
  95\% confidence interval from each model.
\end{enumerate}

\subsection{Analysis plan}\label{analysis-plan}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# The estimand is the treatment effect at the LAST visit, expressed}
\CommentTok{\# as beta\_trt + beta\_\{trt:visit\_f\_last\}. Earlier versions picked the}
\CommentTok{\# \textasciigrave{}trt\textasciigrave{} main effect alone, which under the \textasciigrave{}trt * visit\_f\textasciigrave{}}
\CommentTok{\# parameterisation corresponds to the effect at the REFERENCE visit,}
\CommentTok{\# not the final visit. See docs/morris{-}audit{-}2026{-}04{-}17.md.}
\NormalTok{fit\_mmrm }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(dat) \{}
\NormalTok{  dat }\OtherTok{\textless{}{-}}\NormalTok{ dat }\SpecialCharTok{|\textgreater{}}
    \FunctionTok{group\_by}\NormalTok{(id) }\SpecialCharTok{|\textgreater{}}
    \FunctionTok{mutate}\NormalTok{(}
      \AttributeTok{baseline =} \FunctionTok{first}\NormalTok{(y\_star),}
      \AttributeTok{change =}\NormalTok{ y\_star }\SpecialCharTok{{-}}\NormalTok{ baseline}
\NormalTok{    ) }\SpecialCharTok{|\textgreater{}}
    \FunctionTok{ungroup}\NormalTok{() }\SpecialCharTok{|\textgreater{}}
    \FunctionTok{filter}\NormalTok{(visit }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{) }\SpecialCharTok{|\textgreater{}}
    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{visit\_f =} \FunctionTok{factor}\NormalTok{(visit))}

\NormalTok{  mod }\OtherTok{\textless{}{-}}\NormalTok{ nlme}\SpecialCharTok{::}\FunctionTok{gls}\NormalTok{(}
\NormalTok{    change }\SpecialCharTok{\textasciitilde{}}\NormalTok{ trt }\SpecialCharTok{*}\NormalTok{ visit\_f }\SpecialCharTok{+}\NormalTok{ baseline,}
    \AttributeTok{data =}\NormalTok{ dat,}
    \AttributeTok{correlation =}\NormalTok{ nlme}\SpecialCharTok{::}\FunctionTok{corSymm}\NormalTok{(}\AttributeTok{form =} \SpecialCharTok{\textasciitilde{}}\NormalTok{ visit }\SpecialCharTok{|}\NormalTok{ id),}
    \AttributeTok{weights =}\NormalTok{ nlme}\SpecialCharTok{::}\FunctionTok{varIdent}\NormalTok{(}\AttributeTok{form =} \SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{|}\NormalTok{ visit\_f),}
    \AttributeTok{na.action =}\NormalTok{ na.omit}
\NormalTok{  )}
\NormalTok{  tt }\OtherTok{\textless{}{-}} \FunctionTok{summary}\NormalTok{(mod)}\SpecialCharTok{$}\NormalTok{tTable}
\NormalTok{  last\_visit\_level }\OtherTok{\textless{}{-}} \FunctionTok{max}\NormalTok{(}\FunctionTok{levels}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{visit\_f))}
\NormalTok{  L }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{, }\FunctionTok{nrow}\NormalTok{(tt))}
  \FunctionTok{names}\NormalTok{(L) }\OtherTok{\textless{}{-}} \FunctionTok{rownames}\NormalTok{(tt)}
\NormalTok{  L[}\StringTok{"trt"}\NormalTok{] }\OtherTok{\textless{}{-}} \DecValTok{1}
\NormalTok{  L[}\FunctionTok{paste0}\NormalTok{(}\StringTok{"trt:visit\_f"}\NormalTok{, last\_visit\_level)] }\OtherTok{\textless{}{-}} \DecValTok{1}
\NormalTok{  est }\OtherTok{\textless{}{-}} \FunctionTok{sum}\NormalTok{(L }\SpecialCharTok{*}\NormalTok{ tt[, }\StringTok{"Value"}\NormalTok{])}
\NormalTok{  se }\OtherTok{\textless{}{-}} \FunctionTok{sqrt}\NormalTok{(}\FunctionTok{as.numeric}\NormalTok{(}\FunctionTok{t}\NormalTok{(L) }\SpecialCharTok{\%*\%} \FunctionTok{vcov}\NormalTok{(mod) }\SpecialCharTok{\%*\%}\NormalTok{ L))}
\NormalTok{  p }\OtherTok{\textless{}{-}} \DecValTok{2} \SpecialCharTok{*} \FunctionTok{pnorm}\NormalTok{(}\SpecialCharTok{{-}}\FunctionTok{abs}\NormalTok{(est }\SpecialCharTok{/}\NormalTok{ se))}
  \FunctionTok{c}\NormalTok{(}\AttributeTok{est =}\NormalTok{ est, }\AttributeTok{se =}\NormalTok{ se, }\AttributeTok{p =}\NormalTok{ p)}
\NormalTok{\}}

\NormalTok{fit\_cox }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(dat) \{}
\NormalTok{  surv\_dat }\OtherTok{\textless{}{-}}\NormalTok{ dat }\SpecialCharTok{|\textgreater{}}
    \FunctionTok{group\_by}\NormalTok{(id) }\SpecialCharTok{|\textgreater{}}
    \FunctionTok{summarise}\NormalTok{(}
      \AttributeTok{trt =} \FunctionTok{first}\NormalTok{(trt),}
      \AttributeTok{event\_time =}\NormalTok{ \{}
\NormalTok{        conv }\OtherTok{\textless{}{-}} \FunctionTok{which}\NormalTok{(cdr }\SpecialCharTok{\textgreater{}=} \DecValTok{1}\NormalTok{)}
        \ControlFlowTok{if}\NormalTok{ (}\FunctionTok{length}\NormalTok{(conv) }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{) time[conv[}\DecValTok{1}\NormalTok{]] }\ControlFlowTok{else} \FunctionTok{max}\NormalTok{(time)}
\NormalTok{      \},}
      \AttributeTok{event =} \FunctionTok{as.numeric}\NormalTok{(}\FunctionTok{any}\NormalTok{(cdr }\SpecialCharTok{\textgreater{}=} \DecValTok{1}\NormalTok{)),}
      \AttributeTok{.groups =} \StringTok{"drop"}
\NormalTok{    )}
\NormalTok{  mod }\OtherTok{\textless{}{-}} \FunctionTok{coxph}\NormalTok{(}
    \FunctionTok{Surv}\NormalTok{(event\_time, event) }\SpecialCharTok{\textasciitilde{}}\NormalTok{ trt,}
    \AttributeTok{data =}\NormalTok{ surv\_dat}
\NormalTok{  )}
\NormalTok{  s }\OtherTok{\textless{}{-}} \FunctionTok{summary}\NormalTok{(mod)}
  \FunctionTok{c}\NormalTok{(}
    \AttributeTok{est =} \FunctionTok{coef}\NormalTok{(mod),}
    \AttributeTok{se =}\NormalTok{ s}\SpecialCharTok{$}\NormalTok{coefficients[, }\StringTok{"se(coef)"}\NormalTok{],}
    \AttributeTok{p =}\NormalTok{ s}\SpecialCharTok{$}\NormalTok{coefficients[, }\StringTok{"Pr(\textgreater{}|z|)"}\NormalTok{]}
\NormalTok{  )}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\subsection{Performance metrics}\label{performance-metrics}

For each method, we compute:

\begin{itemize}
\tightlist
\item
  \textbf{Bias:} Mean estimate minus true parameter value
\item
  \textbf{Empirical SE:} Standard deviation of estimates across
  replications
\item
  \textbf{Mean model-based SE:} Mean of estimated standard errors
\item
  \textbf{Power:} Proportion of replications with \(p < 0.05\)
\item
  \textbf{Coverage:} Proportion of 95\% CIs containing the true value
\item
  \textbf{Convergence rate:} Proportion of replications producing valid
  estimates
\end{itemize}

\section{Results}\label{results}

\subsection{Summary statistics}\label{summary-statistics}

\begin{table}[!h]
\centering
\caption{\label{tab:summary-table}Simulation results: MMRM vs Cox regression (2000 replications). Monte Carlo SEs per Morris et al. (2019) Table 6.}
\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{lrrrrrrrrrrr}
\toprule
method & n\_valid & bias & mcse\_bias & emp\_se & mcse\_emp\_se & mean\_se & power & mcse\_power & coverage & mcse\_coverage & convergence\\
\midrule
MMRM & 2000 & -0.003 & 0.001 & 0.034 & 0.001 & 0.035 & 0.994 & 0.002 & 0.954 & 0.005 & 1\\
Cox & 2000 & -0.002 & 0.002 & 0.110 & 0.002 & 0.116 & 0.716 & 0.010 & 0.961 & 0.004 & 1\\
\bottomrule
\end{tabular}}
\end{table}

\subsection{Figures}\label{figures}

\section{Discussion}\label{discussion}

The historical review reveals that the convergence of AD trials on MMRM
and change-score estimands resulted from three reinforcing forces: (1)
the regulatory requirement for continuous cognitive and functional
co-primary endpoints, established
by\textsuperscript{\citeproc{ref-leber1990guidelines}{5}}; (2) the
decisive empirical evidence against LOCF as a primary analysis method,
led by\textsuperscript{\citeproc{ref-mallinckrodt2001dropout}{9}} and
codified by the\textsuperscript{\citeproc{ref-nrc2010missing}{14}}
panel; and (3) the statistical efficiency advantages of modeling
repeated continuous data relative to dichotomizing into time-to-event
endpoints, as demonstrated
by\textsuperscript{\citeproc{ref-donohue2011relative}{19}}
and\textsuperscript{\citeproc{ref-li2019relative}{20}}.

Oncology, by contrast, arrived at a different equilibrium because its
primary endpoint---death---is inherently binary and irreversible,
because the proportional hazards framework has deep regulatory
precedent, and because no single continuous severity measure plays the
role that ADAS-Cog or CDR-SB plays in
AD\textsuperscript{\citeproc{ref-fda2018oncology}{1},\citeproc{ref-fleming1996surrogate}{2}}.

The simulation results are expected to confirm the theoretical
predictions from the literature: MMRM should yield substantially greater
power than Cox regression when both are applied to CDR data from the
same simulated trial, because the continuous analysis uses information
from the full trajectory of scores at every visit, whereas the survival
analysis reduces this trajectory to a single binary event. The power
advantage should be most pronounced in the MCI-to-mild-dementia
transition modeled here, where conversion rates are modest and many
participants in a 24-month trial will not reach the CDR \(\geq\) 1.0
threshold, resulting in heavy right-censoring that limits the
information available to the Cox model.

Several caveats warrant emphasis. First, the comparison is inherently
asymmetric: the MMRM uses more information from each participant (all
visits) than the Cox model (only the first conversion event). Second,
the survival endpoint is limited by the visit schedule; true conversion
may occur between visits, introducing interval censoring that the
standard Cox model does not accommodate. Third, the relative efficiency
depends on the threshold chosen; a threshold that most participants
cross during follow-up would shift the balance toward the survival
analysis. Fourth, the clinical relevance of the two estimands differs: a
treatment effect expressed as slowing of CDR-SB decline addresses a
different clinical question than one expressed as delay to a diagnostic
milestone. Both may be meaningful to patients and regulators, and the
choice between them should be driven by the estimand framework
of\textsuperscript{\citeproc{ref-ich2019e9r1}{15}} rather than by
statistical convenience alone.

\section{Future research}\label{future-research}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Additional covariance structures.} The present simulation
  assumes MAR dropout. Extensions under MNAR mechanisms (e.g.,
  pattern-mixture models, selection models) would illuminate the
  robustness of each method under more adversarial missingness.
\item
  \textbf{Nonlinear trajectories.} AD progression is not strictly
  linear, particularly in the transition from MCI to dementia.
  Simulation of sigmoidal or piecewise trajectories would test the
  sensitivity of MMRM (which makes no linearity assumption) relative to
  slope-based alternatives.
\item
  \textbf{Alternative survival formulations.} Interval-censored survival
  models and multi-state models that accommodate transitions among CDR
  stages (0.5 \(\to\) 1 \(\to\) 2 \(\to\) 3) could recover some of the
  information lost by standard Cox regression.
\item
  \textbf{Joint modeling.} Joint models of longitudinal CDR trajectories
  and the time-to-conversion event could unify the two analytic
  approaches, potentially offering both the efficiency of longitudinal
  modeling and the clinical interpretability of a time-to-event
  estimand.
\item
  \textbf{Software tools.} An R package implementing the simulation
  framework and providing power calculation utilities for both MMRM and
  survival designs in AD trials would facilitate trial planning.
\item
  \textbf{Applied validation.} Retrospective application of both methods
  to completed trial datasets (e.g., from the ADCS or ADNI) would
  ground-truth the simulation findings in real data with genuine
  missingness patterns and measurement properties.
\item
  \textbf{Estimand alignment.} Further work is needed to clarify how the
  ICH E9(R1) estimand framework should be applied when the same
  underlying data support both change-score and time-to-event estimands,
  particularly regarding the handling of intercurrent events such as
  treatment switching or rescue medication.
\end{enumerate}

\section{References}\label{references}

\subsection{Morris et al.~(2019) ADEMP
Compliance}\label{morris-et-al.-2019-ademp-compliance}

This simulation study was audited against the reporting standards
proposed by Morris, White, and
Crowther\textsuperscript{\citeproc{ref-morris2019using}{23}}. The full
audit is at \texttt{docs/morris-audit-2026-04-17.md}.

\textbf{Verdict:} Not compliant.

\textbf{Key gaps identified:}

\begin{itemize}
\tightlist
\item
  Most report chunks are \texttt{eval=FALSE}; results are never produced
  by the rendered report.
\item
  Estimand extractor at \texttt{sim\_study.R:92} falls back to
  \texttt{grep(\textquotesingle{}trt\textquotesingle{}){[}1{]}} which
  can silently pick a \texttt{trt:visit\_f} interaction term.
\item
  No Monte Carlo SE on any metric.
\end{itemize}

A remediation plan is documented in the audit file and will be executed
in a subsequent revision.

\protect\phantomsection\label{refs}
\begin{CSLReferences}{0}{1}
\bibitem[\citeproctext]{ref-fda2018oncology}
\CSLLeftMargin{1. }%
\CSLRightInline{U.S. Food and Drug Administration. \emph{Clinical Trial
Endpoints for the Approval of Cancer Drugs and Biologics. Guidance for
Industry}. Center for Drug Evaluation; Research; 2018.}

\bibitem[\citeproctext]{ref-fleming1996surrogate}
\CSLLeftMargin{2. }%
\CSLRightInline{Fleming TR, DeMets DL. Surrogate end points in clinical
trials: Are we being misled? \emph{Annals of Internal Medicine}.
1996;125(7):605-613.
doi:\href{https://doi.org/10.7326/0003-4819-125-7-199610010-00011}{10.7326/0003-4819-125-7-199610010-00011}}

\bibitem[\citeproctext]{ref-donohue2012mmrm}
\CSLLeftMargin{3. }%
\CSLRightInline{Donohue MC, Aisen PS. Mixed model of repeated measures
versus slope models in {Alzheimer's} disease clinical trials. \emph{The
Journal of Nutrition, Health \& Aging}. 2012;16(4):360-364.
doi:\href{https://doi.org/10.1007/s12603-012-0047-7}{10.1007/s12603-012-0047-7}}

\bibitem[\citeproctext]{ref-mallinckrodt2003assessing}
\CSLLeftMargin{4. }%
\CSLRightInline{Mallinckrodt CH, Sanger TM, Dubé S, et al. Assessing and
interpreting treatment effects in longitudinal clinical trials with
missing data. \emph{Biological Psychiatry}. 2003;53(8):754-760.
doi:\href{https://doi.org/10.1016/S0006-3223(02)01867-X}{10.1016/S0006-3223(02)01867-X}}

\bibitem[\citeproctext]{ref-leber1990guidelines}
\CSLLeftMargin{5. }%
\CSLRightInline{Leber P. \emph{Guidelines for the Clinical Evaluation of
Antidementia Drugs, First Draft}. U.S. Food; Drug Administration; 1990.}

\bibitem[\citeproctext]{ref-rosen1984adas}
\CSLLeftMargin{6. }%
\CSLRightInline{Rosen WG, Mohs RC, Davis KL. A new rating scale for
{Alzheimer's} disease. \emph{The American Journal of Psychiatry}.
1984;141(11):1356-1364.
doi:\href{https://doi.org/10.1176/ajp.141.11.1356}{10.1176/ajp.141.11.1356}}

\bibitem[\citeproctext]{ref-morris1993cdr}
\CSLLeftMargin{7. }%
\CSLRightInline{Morris JC. The {Clinical Dementia Rating} ({CDR}):
Current version and scoring rules. \emph{Neurology}.
1993;43(11):2412-2414.
doi:\href{https://doi.org/10.1212/WNL.43.11.2412-a}{10.1212/WNL.43.11.2412-a}}

\bibitem[\citeproctext]{ref-fda2018earlyad}
\CSLLeftMargin{8. }%
\CSLRightInline{U.S. Food and Drug Administration. \emph{Early
{Alzheimer's} Disease: Developing Drugs for Treatment. Guidance for
Industry}. Center for Drug Evaluation; Research; 2018.}

\bibitem[\citeproctext]{ref-mallinckrodt2001dropout}
\CSLLeftMargin{9. }%
\CSLRightInline{Mallinckrodt CH, Clark WS, David SR. Accounting for
dropout bias using mixed-effects models. \emph{Journal of
Biopharmaceutical Statistics}. 2001;11(1--2):9-21.
doi:\href{https://doi.org/10.1081/BIP-100104194}{10.1081/BIP-100104194}}

\bibitem[\citeproctext]{ref-mallinckrodt2001typei}
\CSLLeftMargin{10. }%
\CSLRightInline{Mallinckrodt CH, Clark WS, David SR. Type {I} error
rates from mixed effects model repeated measures versus fixed effects
{ANOVA} with missing values imputed via last observation carried
forward. \emph{Drug Information Journal}. 2001;35(4):1215-1225.
doi:\href{https://doi.org/10.1177/009286150103500418}{10.1177/009286150103500418}}

\bibitem[\citeproctext]{ref-mallinckrodt2003response}
\CSLLeftMargin{11. }%
\CSLRightInline{Mallinckrodt CH, Clark WS, Carroll RJ, Molenberghs G.
Assessing response profiles from incomplete longitudinal clinical trial
data under regulatory considerations. \emph{Journal of Biopharmaceutical
Statistics}. 2003;13(2):179-190.
doi:\href{https://doi.org/10.1081/BIP-120019265}{10.1081/BIP-120019265}}

\bibitem[\citeproctext]{ref-obrien2005semiparametric}
\CSLLeftMargin{12. }%
\CSLRightInline{O'Brien PC, Zhang D, Bailey KR. Semi-parametric and
non-parametric methods for clinical trials with incomplete data.
\emph{Statistics in Medicine}. 2005;24(3):341-358.
doi:\href{https://doi.org/10.1002/sim.1963}{10.1002/sim.1963}}

\bibitem[\citeproctext]{ref-molenberghs2004analyzing}
\CSLLeftMargin{13. }%
\CSLRightInline{Molenberghs G, Thijs H, Jansen I, et al. Analyzing
incomplete longitudinal clinical trial data. \emph{Biostatistics}.
2004;5(3):445-464.
doi:\href{https://doi.org/10.1093/biostatistics/kxh001}{10.1093/biostatistics/kxh001}}

\bibitem[\citeproctext]{ref-nrc2010missing}
\CSLLeftMargin{14. }%
\CSLRightInline{National Research Council. \emph{The Prevention and
Treatment of Missing Data in Clinical Trials}. The National Academies
Press; 2010. doi:\href{https://doi.org/10.17226/12955}{10.17226/12955}}

\bibitem[\citeproctext]{ref-ich2019e9r1}
\CSLLeftMargin{15. }%
\CSLRightInline{International Council for Harmonisation. \emph{{ICH
E9(R1)}: Addendum on Estimands and Sensitivity Analysis in Clinical
Trials to the Guideline on Statistical Principles for Clinical Trials}.
ICH; 2019.}

\bibitem[\citeproctext]{ref-wang2018dpm}
\CSLLeftMargin{16. }%
\CSLRightInline{Wang G, Berry S, Xiong C, et al. A novel cognitive
disease progression model for clinical trials in autosomal-dominant
{Alzheimer's} disease. \emph{Statistics in Medicine}.
2018;37(21):3047-3055.
doi:\href{https://doi.org/10.1002/sim.7811}{10.1002/sim.7811}}

\bibitem[\citeproctext]{ref-buddhaeberlein2022aducanumab}
\CSLLeftMargin{17. }%
\CSLRightInline{Budd Haeberlein S, Aisen PS, Barkhof F, et al. Two
randomized phase 3 studies of aducanumab in early {Alzheimer's} disease.
\emph{The Journal of Prevention of Alzheimer's Disease}.
2022;9(2):197-210.
doi:\href{https://doi.org/10.14283/jpad.2022.30}{10.14283/jpad.2022.30}}

\bibitem[\citeproctext]{ref-dhadda2022lecanemab}
\CSLLeftMargin{18. }%
\CSLRightInline{Dhadda S, Kanekiyo M, Li D, et al. Consistency of
efficacy results across various clinical measures and statistical
methods in the lecanemab phase 2 trial of early {Alzheimer's} disease.
\emph{Alzheimer's Research \& Therapy}. 2022;14(1):182.
doi:\href{https://doi.org/10.1186/s13195-022-01129-x}{10.1186/s13195-022-01129-x}}

\bibitem[\citeproctext]{ref-donohue2011relative}
\CSLLeftMargin{19. }%
\CSLRightInline{Donohue MC, Gamst AC, Thomas RG, et al. The relative
efficiency of time-to-threshold and rate of change in longitudinal data.
\emph{Contemporary Clinical Trials}. 2011;32(5):685-693.
doi:\href{https://doi.org/10.1016/j.cct.2011.04.007}{10.1016/j.cct.2011.04.007}}

\bibitem[\citeproctext]{ref-li2019relative}
\CSLLeftMargin{20. }%
\CSLRightInline{Li D, Iddi S, Aisen PS, Thompson WK, Donohue MC. The
relative efficiency of time-to-progression and continuous measures of
cognition in presymptomatic {Alzheimer's} disease. \emph{Alzheimer's \&
Dementia: Translational Research \& Clinical Interventions}.
2019;5:308-318.
doi:\href{https://doi.org/10.1016/j.trci.2019.04.004}{10.1016/j.trci.2019.04.004}}

\bibitem[\citeproctext]{ref-jonsson2024progression}
\CSLLeftMargin{21. }%
\CSLRightInline{Jönsson L, Ivkovic M, Atri A, et al. Progression
analysis versus traditional methods to quantify slowing of disease
progression in {Alzheimer's} disease. \emph{Alzheimer's Research \&
Therapy}. 2024;16(1):48.
doi:\href{https://doi.org/10.1186/s13195-024-01413-y}{10.1186/s13195-024-01413-y}}

\bibitem[\citeproctext]{ref-ito2014predicting}
\CSLLeftMargin{22. }%
\CSLRightInline{Ito K, Hutmacher MM. Predicting the time to clinically
worsening in mild cognitive impairment patients and its utility in
clinical trial design by modeling a longitudinal {Clinical Dementia
Rating Sum of Boxes} from the {ADNI} database. \emph{Journal of
Alzheimer's Disease}. 2014;40(4):967-979.
doi:\href{https://doi.org/10.3233/JAD-132090}{10.3233/JAD-132090}}

\bibitem[\citeproctext]{ref-morris2019using}
\CSLLeftMargin{23. }%
\CSLRightInline{Morris TP, White IR, Crowther MJ. Using simulation
studies to evaluate statistical methods. \emph{Statistics in Medicine}.
2019;38(11):2074-2102.
doi:\href{https://doi.org/10.1002/sim.8086}{10.1002/sim.8086}}

\end{CSLReferences}

\end{document}
